Tactic Script Optimisation for Aesop

### 2024-09-09 paper.tex
Rewrite equality section

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage{csquotes}
\usepackage{subcaption}
\usepackage{microtype}
\usepackage{newunicodechar}
\usepackage[plain]{algorithm}
\usepackage{algpseudocodex}

@@

\newunicodechar{₁}{\ensuremath{_{1}}}
\newunicodechar{₂}{\ensuremath{_{2}}}
\newunicodechar{₃}{\ensuremath{_{3}}}
\newunicodechar{₄}{\ensuremath{_{4}}}
\newunicodechar{₅}{\ensuremath{_{5}}}
\newunicodechar{₆}{\ensuremath{_{6}}}
\newunicodechar{₇}{\ensuremath{_{7}}}
\newunicodechar{₈}{\ensuremath{_{8}}}
\newunicodechar{₉}{\ensuremath{_{9}}}
\newunicodechar{ₙ}{\ensuremath{_{n}}}
\newunicodechar{ₘ}{\ensuremath{_{m}}}
\newunicodechar{ᵢ}{\ensuremath{_{i}}}
\newunicodechar{ⱼ}{\ensuremath{_{j}}}
\newunicodechar{ₗ}{\ensuremath{_{l}}}
\newunicodechar{ₖ}{\ensuremath{_{k}}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{Δ}{\ensuremath{\Delta}}
\newunicodechar{Φ}{\ensuremath{\Phi}}
\newunicodechar{Ψ}{\ensuremath{\Psi}}
\newunicodechar{Ω}{\ensuremath{\Omega}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{∉}{\ensuremath{\notin}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{≢}{\ensuremath{\nequiv}}
\newunicodechar{≠}{\ensuremath{\neq}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≔}{\ensuremath{\coloneq}}
\newunicodechar{∧}{\ensuremath{\land}}
\newunicodechar{∨}{\ensuremath{\lor}}
\newunicodechar{→}{\ensuremath{\rightarrow}}
\newunicodechar{↔}{\ensuremath{\leftrightarrow}}
\newunicodechar{⊢}{\ensuremath{\vdash}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{∷}{\ensuremath{\mathbin{::}}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}

\newcommand{\tac}[1]{\ensuremath{\mathrm{\mathcolor{blue}{#1}}}}
\newcommand{\Lam}[2]{λ\,#1,\; #2}
\newcommand{\mvar}[1]{{?#1}}
\newcommand{\pre}{\mathrm{pre}}
\newcommand{\post}{\mathrm{post}}
\newcommand{\tacbullet}{\vcenter{\hbox{\scalebox{.7}{$\bullet\;\;$}}}}
\newcommand{\step}[5]{\ensuremath{(#1,\allowbreak #2,\allowbreak #3,\allowbreak #4,\allowbreak #5)}}

@@

Before we can start optimising the script, we need one piece of infrastructure.
Many of our optimisations will need to check whether two tactics $t₁$ and $t₂$, when run in the same tactic state $(Φ, G)$, produce the same result.
This is non-trivial because even when we run the same tactic twice, the internal names of any resulting metavariables and fvars, which are produced by a global name generator, change.
So we need a procedure that checks whether the results of $t₁$ and $t₂$ -- two tactic states $(Φ₁, G₁)$ and $(Φ₂, G₂)$ -- are equal up to internal names.
This procedure is implemented in file \texttt{Aesop/Util/EqualUpToIds.lean}.

% TODO precisely specify the equality: it's structural equality up to renaming
More precisely, the procedure checks for a variant of \emph{structural equality} which we call $≡$.
Since $≡$ is structural, different but definitionally equal expressions are considered different: $(\Lam{x : A}{x})~a ≢ a$.
Moreover, names of bound variables matter, so α-equivalent terms are not identified: $\Lam{x : A}{x} ≢ \Lam{y : A}{y}$.
We choose such a strict notion of equality because it is not unreasonable for tactics to distinguish between definitionally equal or α-equivalent expressions.
However, the procedure ignores Lean's special \enquote{implementation detail} hypotheses, which are invisible to users and do not affect most tactics.

To implement this notion of equality, we maintain a global bijective map $M$ from (internal names of) unassigned metavariables in $Φ₁$ to (internal names of) unassigned metavariables in $Φ₂$.
$M$ maps $\mvar{m₁}$ to $\mvar{m₂}$ if (1) $\mvar{m₁}$ and $\mvar{m₂}$ have equal local contexts and types and (2) we have previously encountered $\mvar{m₁}$ and $\mvar{m₂}$ in the same position of some expression.
For example, when comparing expressions $f~\mvar{m₁}~\mvar{m₂}$ and $f~\mvar{m₃}~\mvar{m₄}$, we update $M$ with the mappings $M(\mvar{m₁}) ≔ \mvar{m₃}$ and $M(\mvar{m₂}) ≔ \mvar{m₄}$.
When we later encounter $\mvar{m₁}$ again, we consider it equal to an expression $e₂$ if and only if $e₂ = M(\mvar{m₁})$.
This scheme correctly identifies $f~\mvar{m₁}~\mvar{m₁}$ as equal to $f~\mvar{m₂}~\mvar{m₂}$ but different from $f~\mvar{m₂}~\mvar{m₃}$ (supposing that $\mvar{m₁}$, $\mvar{m₂}$ and $\mvar{m₃}$ all have equal local contexts and types).

We now consider four mutually recursive procedures that compare expressions, local contexts, unassigned metavariables and tactic states.

\paragraph{Tactic states} To compare $(Φ₁, G₁)$ and $(Φ₂, G₂)$, we first check whether $G₁$ and $G₂$ contain the same number of goals.
If so, we compare each goal (unassigned metavariable) $\mvar{m₁}$ in $G₁$ to its counterpart $\mvar{m₂}$ in $G₂$, using $Φ₁$ as the metavariable context for $\mvar{m₁}$ and $Φ₂$ as the metavariable context for $\mvar{m₂}$.

\paragraph{Unassigned metavariables} To compare $\mvar{m₁}$ in context $Φ₁$ and $\mvar{m₂}$ in $Φ₂$, we first check whether $M(\mvar{m₁})$ is defined.
If so, $\mvar{m₁}$ and $\mvar{m₂}$ are equal exactly if $M(\mvar{m₁}) = \mvar{m₂}$.
If not, we first compare the local contexts $σ₁$ of $\mvar{m₁}$ (in $Φ₁$) and $σ₂$ of $\mvar{m₂}$ (in $Φ₂$).
If these are equal, we compare the type $T₁$ of $\mvar{m₁}$ (in $Φ₁$ and with local context $σ₁$) to the type $T₂$ of $\mvar{m₂}$ (in $Φ₂$ and $σ₂$).

\paragraph{Local contexts} To compare a local context $σ₁$ in metavariable context $Φ₁$ with $σ₂$ in $Φ₂$, we first check whether they contain an equal number of fvars.
If so, we check whether each fvar $h₁ : T₁$ in $σ₁$ is equal to the fvar $h₂ : T₂$ at the same position in $σ₂$; that is, whether $T₁ ≡ T₂$.
If $h₁$ has an associated value ($h₁ : T₁ ≔ v₁$), then $h₂$ must additionally have a value $v₂$ with $v₁ ≡ v₂$.

However, fvars can also appear in other expressions, so once we have confirmed that $h₁$ and $h₂$ are equal, we should treat them as equal despite their different names.
To that end, we employ a similar scheme as for metavariables: while iterating through the local contexts, we maintain a bijective map $F$ from fvars in $σ₁$ to fvars in $σ₂$.
Once we have determined $h₁ ≡ h₂$, we add the mapping $F(h₁) ≔ h₂$.
In subsequent fvars and in the types of the metavariables whose local contexts we are processing, we then consider $h₁$ equal to an expression $e₂$ exactly if $e₂ = h₂$.

\paragraph{Expressions} To compare an expression $e₁$ in metavariable context $Φ₁$ and local context $σ₁$ with an expression $e₂$ in $Φ₂$ and $σ₂$, we perform the expected structural comparison.
For example, $e₁ = \Lam{x₁ : A₁}{b₁}$ and $e₂ = \Lam{x₂ : A₂}{b₂}$ are equal if $x₁ = x₂$, $A₁ ≡ A₂$ and $b₁ ≡ b₂$.

The only expressions requiring special treatment are metavariables.
If $e₁ = \mvar{m₁}$ and $e₂ = \mvar{m₂}$ and both $\mvar{m₁}$ and $\mvar{m₂}$ are unassigned, then we compare $\mvar{m₁}$ and $\mvar{m₂}$ as unassigned metavariables.
If both $\mvar{m₁}$ and $\mvar{m₂}$ are assigned, we compare their assignments.
Otherwise $\mvar{m₁}$ and $\mvar{m₂}$ are unequal.
(Lean metavariables can also be \enquote{delayed assigned}, but I ignore this possibility here for clarity.)

@@

For the dynamic reordering algorithm of Section~\ref{sec:dynamic-reordering}, we need a slightly relaxed equality $≡_{a}$.
In this variant, we allow differences in whether a metavariable is assigned, so an unassigned metavariable $\mvar{m₁}$ can be $≡_{a}$-equal to an arbitrary (type-correct) expression $e₂$.
However, if $\mvar{m₁}$ appears multiple times, then the corresponding expressions $e₂$, $e₂'$, etc.\ must all be $≡_{a}$-equal.
In other words, we optimistically assume that an expression $e₁$ with $e₁ ≡_{a} e₂$ will eventually be assigned to $\mvar{m₁}$ by a later tactic, but we ensure that such an assignment is possible by requiring $e₂ ≡_{a} e₂' ≡_{a} \dots$.
Symmetrically, an unassigned metavariable $\mvar{m₂}$ from the right-hand expression can be equal to an arbitrary left-hand expression $e₁$ under analogous conditions.

To implement this check, we extend the state of our procedures with two maps $A_{l}$ and $A_{r}$.
$A_{l}$ maps each unassigned metavariable from a left-hand expression to the first corresponding right-hand expression; $A_{r}$ does the same for right-hand unassigned metavariables.

@@

Applying the steps is somewhat more subtle than in the static variant of the algorithm, so the function \textsc{apply-step} is more complex.
We first run the step's tactic $t$ on the current goal $g$ in the current metavariable context $Φ$.
If the tactic fails, \textsc{apply-step} implicitly also fails.
If the tactic succeeds, we check whether the produced goals $G'$ (in the produced context $Ψ$) are equal to the expected goals $G^{\post}$ of the step (in the step's post-context $Φ^{\post}$).
Here we use the equality from Section~\ref{sec:equality}: the goals in $G'$ should be equal to the goals in $G^{\post}$ up to names of fvars and metavariables.
This is needed since different runs of the same tactic (even in the same metavariable context) will almost always generate different names.
Furthermore, we use the variant $≡_{a}$, so we allow differences in whether a metavariable is assigned.
This is needed to account for the expected effects of reordering: if a tactic $t₁$ assigns a metavariable and a tactic $t₂$ was run after $t₁$ in the original order, but now $t₂$ is run before $t₁$, then we cannot expect the metavariable to be assigned after $t₂$.
If the goals are not equal in this sense, \textsc{apply-step} again fails.

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage{csquotes}
\usepackage{subcaption}
\usepackage{microtype}
\usepackage{newunicodechar}
\usepackage[plain]{algorithm}
\usepackage{algpseudocodex}
\usepackage{mathtools}

@@

\newunicodechar{₁}{\ensuremath{_{1}}}
\newunicodechar{₂}{\ensuremath{_{2}}}
\newunicodechar{₃}{\ensuremath{_{3}}}
\newunicodechar{₄}{\ensuremath{_{4}}}
\newunicodechar{₅}{\ensuremath{_{5}}}
\newunicodechar{₆}{\ensuremath{_{6}}}
\newunicodechar{₇}{\ensuremath{_{7}}}
\newunicodechar{₈}{\ensuremath{_{8}}}
\newunicodechar{₉}{\ensuremath{_{9}}}
\newunicodechar{ₙ}{\ensuremath{_{n}}}
\newunicodechar{ₘ}{\ensuremath{_{m}}}
\newunicodechar{ᵢ}{\ensuremath{_{i}}}
\newunicodechar{ⱼ}{\ensuremath{_{j}}}
\newunicodechar{ₗ}{\ensuremath{_{l}}}
\newunicodechar{ₖ}{\ensuremath{_{k}}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{Δ}{\ensuremath{\Delta}}
\newunicodechar{Φ}{\ensuremath{\Phi}}
\newunicodechar{Ψ}{\ensuremath{\Psi}}
\newunicodechar{Ω}{\ensuremath{\Omega}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{∉}{\ensuremath{\notin}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{≢}{\ensuremath{\nequiv}}
\newunicodechar{≠}{\ensuremath{\neq}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≔}{\ensuremath{\coloneq}}
\newunicodechar{∧}{\ensuremath{\land}}
\newunicodechar{∨}{\ensuremath{\lor}}
\newunicodechar{∪}{\ensuremath{\cup}}
\newunicodechar{∩}{\ensuremath{\cap}}
\newunicodechar{→}{\ensuremath{\rightarrow}}
\newunicodechar{↔}{\ensuremath{\leftrightarrow}}
\newunicodechar{↦}{\ensuremath{\mapsto}}
\newunicodechar{⊢}{\ensuremath{\vdash}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{∷}{\ensuremath{\mathbin{::}}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}

\newcommand{\tac}[1]{\ensuremath{\mathrm{\mathcolor{blue}{#1}}}}
\newcommand{\Lam}[2]{λ\,#1,\; #2}
\newcommand{\mvar}[1]{{?#1}}
\newcommand{\pre}{\mathrm{pre}}
\newcommand{\post}{\mathrm{post}}
\newcommand{\tacbullet}{\vcenter{\hbox{\scalebox{.7}{$\bullet\;\;$}}}}
\newcommand{\step}[5]{\ensuremath{(#1,\allowbreak #2,\allowbreak #3,\allowbreak #4,\allowbreak #5)}}
\newcommand{\aeq}{\hat{≡}}
\newcommand{\dom}{\ensuremath{\mathrm{dom}}}

@@

Before we can start optimising the script, we need one piece of infrastructure.
Many of our optimisations will need to check whether two tactics $t₁$ and $t₂$, when run in the same tactic state $(Φ, G)$, produce the same result.
This is non-trivial because even when we run the same tactic twice, the internal names of any resulting metavariables and fvars, which are produced by a global name generator, change.
So we need a procedure that checks whether the results of $t₁$ and $t₂$ -- two tactic states $(Φ₁, G₁)$ and $(Φ₂, G₂)$ -- are equal up to internal names.

More precisely, we say that two expressions $e₁$ and $e₂$ are \emph{structurally equal}, $e₁ = e₂$, if $e₁$ and $e₂$ are the same expression.
In particular, different but definitionally equal expressions are not structurally equal: $(\Lam{x : A}{x})~a ≠ a$.
Moreover, names of bound variables matter: $\Lam{x : A}{x} ≠ \Lam{y : A}{y}$.
This strict notion of equality is a suitable starting point because tactics may distinguish between definitionally equal or α-equivalent expressions.

We then define how to check \emph{structural equality up to renaming}, written $≡$, of two tactic states $(Φ₁, G₁)$ and $(Φ₂, G₂)$.
Throughout this procedure, we maintain a partial map $σ$ from metavariable names to metavariable names.
A mapping $\mvar{m₁} ↦ \mvar{m₂}$ in $σ$ means that the metavariable $\mvar{m₁}$ occurring in $G₁$ (either as a goal or in the type or local context of a goal) is equal up to renaming to the metavariable $\mvar{m₂}$ occurring in $G₂$.

To check $(Φ₁, G₁) ≡ (Φ₂, G₂)$, we iterate through $G₁$ and $G₂$, checking that each goal (unassigned metavariable) $\mvar{m₁}$ from $G₁$ is equal up to renaming to the corresponding goal $\mvar{m₂}$ from $G₂$.
(If $G₁$ and $G₂$ have different lengths, the tactic states cannot be equal.)

To check $\mvar{m₁} ≡ \mvar{m₂}$, where $\mvar{m₁}$ is an unassigned metavariable in $Φ₁$ and $\mvar{m₂}$ is an unassigned metavariable in $Φ₂$, we proceed as follows:
\begin{itemize}
  \item If $\mvar{m₁} = \mvar{m₂}$ (i.e., the metavariables have the same internal name), $\mvar{m₁} ≡ \mvar{m₂}$.
  \item If $σ(\mvar{m₁}) = \mvar{m₂}$, then $\mvar{m₁} ≡ \mvar{m₂}$.
  \item If $\mvar{m₁} ∈ \dom(σ)$, but $σ(\mvar{m₁}) ≠ \mvar{m₂}$, then $\mvar{m₁} ≢ \mvar{m₂}$.
  \item Otherwise we check that the local contexts and types of $\mvar{m₁}$ and $\mvar{m₂}$ are equal up to renaming.
        On success, the mapping $\mvar{m₁} ↦ \mvar{m₂}$ is added to $σ$.
\end{itemize}

To check $Γ₁ ≡ Γ₂$, where $Γ₁$ and $Γ₂$ are local contexts, we check that for each fvar $h₁ : T₁$ in $Γ₁$ and corresponding fvar $h₂ : T₂$ in $Γ₂$, $T₁ ≡ T₂$.
However, we must take into account that the internal names of fvars in $Γ₁$ and $Γ₂$ may differ.
Hence, we maintain another partial map $τ$ from fvar names to fvar names.
A mapping $h₁ ↦ h₂$ in $τ$ means that the fvar $h₁$ in $Γ₁$ corresponds to $h₂$ in $Γ₂$, so once the types of each pair of fvars have been checked, the pair is added to $τ$.

Finally, to check $e₁ ≡ e₂$, where $e₁$ and $e₂$ are expressions (including types), we check that $e₁$ and $e₂$ are structurally equal, with the following exceptions:
\begin{itemize}
  \item Any metavariable from $e₁$ that is assigned in $Φ₁$ is replaced with its assignment before checking for structural equality, and similar for $e₂$ and $Φ₂$.
  \item An fvar $h₁$ in $e₁$ and an fvar $h₂$ at the same position in $e₂$ are considered equal if either $h₁ = h₂$ (i.e., $h₁$ and $h₂$ have the same internal name) or $τ(h₁) = h₂$.
  \item An unassigned metavariable $\mvar{m₁}$ in $e₁$ and an unassigned metavariable $\mvar{m₂}$ at the same position in $e₂$ are considered equal if $\mvar{m₁} ≡ \mvar{m₂}$.
\end{itemize}

For example, when checking $f~\mvar{m₁}~\mvar{m₁} ≡ f~\mvar{m₂}~\mvar{m₂}$ (with $σ$ initially empty), we check whether $\mvar{m₁}$ and $\mvar{m₂}$ have equal local context and type.
If so, the mapping $\mvar{m₁} ↦ \mvar{m₂}$ is added to $σ$.
The check at the second occurrence of $\mvar{m₁}$ and $\mvar{m₂}$ then succeeds immediately because $σ(\mvar{m₁}) = \mvar{m₂}$.
Conversely, when checking $f~\mvar{m₁}~\mvar{m₁} = f~\mvar{m₂}~\mvar{m₃}$, while the check of $\mvar{m₁} ≡ \mvar{m₂}$ succeeds, the check of $\mvar{m₁} ≡ \mvar{m₃}$ fails since at this point $\mvar{m₁} ∈ \dom(σ)$ but $σ(\mvar{m₁}) = m₂ ≠ m₃$.

@@

For the dynamic reordering algorithm of Section~\ref{sec:dynamic-reordering}, we need a slightly relaxed equality $\aeq$.
In this variant, we allow differences in whether a metavariable is assigned, so an unassigned metavariable $\mvar{m₁}$ can be $\aeq$-equal to an arbitrary (type-correct) expression $e₂$.
However, if $\mvar{m₁}$ appears multiple times, then the corresponding expressions $e₂$, $e₂'$, etc.\ must all be $\aeq$-equal.
In other words, we optimistically assume that an expression $e₁$ with $e₁ \aeq e₂$ will be assigned to $\mvar{m₁}$ by a later tactic, but we ensure that such an assignment is possible.
Symmetrically, an unassigned metavariable $\mvar{m₂}$ from the right-hand expression can be equal to an arbitrary left-hand expression $e₁$.

The above description omits some special cases, e.g.\ local definitions and delayed-assigned metavariables.
See the implementation in file \texttt{Aesop/Util/EqualUpToIds.lean} for the full details.

@@

Applying the steps is somewhat more subtle than in the static variant of the algorithm, so the function \textsc{apply-step} is more complex.
We first run the step's tactic $t$ on the current goal $g$ in the current metavariable context $Φ$.
If the tactic fails, \textsc{apply-step} implicitly also fails.
If the tactic succeeds, we check whether the produced goals $G'$ (in the produced context $Ψ$) are equal to the expected goals $G^{\post}$ of the step (in the step's post-context $Φ^{\post}$).
Here we use the equality from Section~\ref{sec:equality}: the goals in $G'$ should be equal to the goals in $G^{\post}$ up to names of fvars and metavariables.
This is needed since different runs of the same tactic (even in the same metavariable context) will almost always generate different names.
Furthermore, we use the variant $\aeq$, so we allow differences in whether a metavariable is assigned.
This is needed to account for the expected effects of reordering: if a tactic $t₁$ assigns a metavariable and a tactic $t₂$ was run after $t₁$ in the original order, but now $t₂$ is run before $t₁$, then we cannot expect the metavariable to be assigned after $t₂$.
If the goals are not equal in this sense, \textsc{apply-step} again fails.
